---
title: "Network Graph Map v1"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---


https://www.r-bloggers.com/2018/05/three-ways-of-visualizing-a-graph-on-a-map/

```{r}
library(assertthat)
library(dplyr)
library(purrr)
library(igraph)
library(ggplot2)
library(ggraph)
library(ggmap)
```

```{r}
country_coords_txt <- "
 1     3.00000  28.00000       Algeria
 2    54.00000  24.00000           UAE
 3   139.75309  35.68536         Japan
 4    45.00000  25.00000 'Saudi Arabia'
 5     9.00000  34.00000       Tunisia
 6     5.75000  52.50000   Netherlands
 7   103.80000   1.36667     Singapore
 8   124.10000  -8.36667         Korea
 9    -2.69531  54.75844            UK
10    34.91155  39.05901        Turkey
11  -113.64258  60.10867        Canada
12    77.00000  20.00000         India
13    25.00000  46.00000       Romania
14   135.00000 -25.00000     Australia
15    10.00000  62.00000        Norway"

# nodes come from the above table and contain geo-coordinates for some
# randomly picked countries
nodes <- read.delim(text = country_coords_txt, header = FALSE,
                    quote = "'", sep = "",
                    col.names = c('id', 'lon', 'lat', 'name'))
```

```{r}
nodes
```

```{r}
# Create connections
set.seed(123) # set random generator state for the same output

N_EDGES_PER_NODE_MIN <-1
N_EDGES_PER_NODE_MAX <- 4
N_CATEGORIES <- 4

# edges: create random connections between countries (nodes)
edges <- map_dfr(nodes$id, function(id) {
  n <- floor(runif(1, N_EDGES_PER_NODE_MIN, N_EDGES_PER_NODE_MAX+1))
  to <- sample(1:max(nodes$id), n, replace=FALSE)
  to <- to[to != id]
  categories <- sample(1:N_CATEGORIES, length(to), replace=TRUE)
  weights <- runif(length(to))
  data_frame(from=id, to=to, weight=weights, category=categories)
  })

edges <- edges %>% mutate(category = as.factor(category))

edges
```

```{r}
g <- graph_from_data_frame(edges, directed=FALSE, vertices=nodes)
```

```{r}
edges_for_plot <- edges %>%
  inner_join(nodes %>% select(id, lon, lat), by=c('from' = 'id')) %>%
  rename(x=lon, y=lat) %>%
  inner_join(nodes %>% select(id, lon, lat), by=c('to' = 'id')) %>%
  rename(xend=lon, yend=lat)

assert_that(nrow(edges_for_plot) == nrow(edges))



```

```{r}
nodes$weight = degree(g)
```

```{r}
maptheme <- theme(panel.grid = element_blank()) +
  theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(axis.title = element_blank()) +
  theme(legend.position = "bottom") +
  theme(panel.grid = element_blank()) +
  theme(panel.background = element_rect(fill="#596673")) +
  theme(plot.margin = unit(c(0,0,0.5,0), 'cm'))
```

```{r}
country_shapes <- geom_polygon(aes(x=long, y=lat, group=group),
                               data=map_data('world'),
                               fill="#CECECE", color="#515151",
                               size=0.15)
mapcoords <- coord_fixed(xlim = c(-150, 180), ylim = c(-55, 80))
```

```{r}
ggplot(nodes) + country_shapes +
  geom_curve(aes(x = x, y = y, xend = xend, yend = yend,     # draw edges as arcs
                 color = category, size = weight),
             data = edges_for_plot, curvature = 0.33,
             alpha = 0.5) +
  scale_size_continuous(guide = FALSE, range = c(0.25, 2)) + # scale for edge widths
  geom_point(aes(x = lon, y = lat, size = weight),           # draw nodes
             shape = 21, fill = 'white',
             color = 'black', stroke = 0.5) +
  scale_size_continuous(guide = "none", range = c(1, 6)) +    # scale for node size
  geom_text(aes(x = lon, y = lat, label = name),             # draw text labels
            hjust = 0, nudge_x = 1, nudge_y = 4,
            size = 3, color = "white", fontface = "bold") +
  mapcoords + maptheme
```

```{r}
ggplot(nodes) + country_shapes +
  geom_curve(aes(x = x, y = y, xend = xend, yend = yend,     # draw edges as arcs
                 color = category, size = weight),
             data = edges_for_plot, curvature = 0.33,
             alpha = 0.5) +
  scale_size_continuous(guide = "none", range = c(0.25, 2)) + # scale for edge widths
  geom_point(aes(x = lon, y = lat),                          # draw nodes
             shape = 21, size = 3, fill = 'white',
             color = 'black', stroke = 0.5) +
  geom_text(aes(x = lon, y = lat, label = name),             # draw text labels
            hjust = 0, nudge_x = 1, nudge_y = 4,
            size = 3, color = "white", fontface = "bold") +
  mapcoords + maptheme
```

# START HERE WITH ATTEMPT OF THE NETWORK GRAPH OF CITI BIKE STATIONS


```{r}
# Let's create some other columns of value
# weekday
# day of the month
# trip duration in minutes
# start hour
citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = wday(ymd_hms(started_at),label=TRUE,abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual))

#citibike
```

Columns
- ride_id: Unique identifier of the ride, probably not meaningful to me
- rideable_type:

```{r}
nrow(citibike)
```

okay good, I've got all the rows from Citibike NYC trips for Sept. 2022

```{r}
# Figure out abandoned and label them
citibike$end_station_name[citibike$end_station_name == ''] <- "Abandoned"
citibike$end_station_id[citibike$end_station_id  == ''] <- "ABAN"
```

this means there are 8012 abandoned bikes if the end_station_id is not present, but the lat/long does exist for more than half

```{r}
citibike %>%
  filter(citibike$end_station_id  == 'ABAN')
```
3838 have NA end_lat and end_lng, and thus abandoned and no final destination



```{r}
citibike %>%
  filter(startsWith(end_station_id, 'SYS'))
```

There are start_station_id and end_station_id that start with 'SYS' with a valid lat/long, so keep them

There are NO starts with JC or HB, but there are ends with JC and HB

```{r}
# Create table of start to end station IDs
stations_cols <- citibike %>%
  select(start_station_id, end_station_id)
stations_table <- as.data.frame((table(stations_cols)))

stations_table <- stations_table %>%
  filter(Freq > 0)

stations_table_order <- stations_table[order(-stations_table$Freq),]

stations_table_order
```

```{r}
stations_table_dif <- stations_table_order %>%
  filter(as.character(start_station_id) != as.character(end_station_id))

nrow(stations_table_dif)
```

471313 rows in which the start and end are different values 

```{r}
# Create table of start to end station IDs
stations_cols <- citibike %>%
  select(start_station_id, end_station_id)
stations_table <- as.data.frame((table(stations_cols)))

stations_table <- stations_table %>%
  filter(Freq > 0)

stations_table_order <- stations_table[order(-stations_table$Freq),]

stations_table_order
```

For all valid combinations of start to end, there are 472920 combinations

The above table also indicates the frequency of those trips for start-end combinations.

The top 5 entries are the same start and end station ID. so these are casual rides, I have a feeling these 5 are near tourist destinations, perhaps Central Park

```{r}
stations_table_dif <- stations_table_order %>%
  filter(as.character(start_station_id) != as.character(end_station_id))

nrow(stations_table_dif)
```

471313 in which the start and the end station IDs are NOT the same

```{r}
tail(stations_table_dif, 200)
```

Now, I want to determine overall flow between any existing combination. At the end of the month, is the flow to or from a station

```{r}
stations_table_dif1 <- stations_table_dif
stations_table_dif2 <- stations_table_dif
```



```{r}
# Added all=TRUE to account for one-sided counts
res <- 
merge(stations_table_dif1, 
      stations_table_dif1, 
      by.x=c('start_station_id','end_station_id'), 
      by.y=c('end_station_id','start_station_id'), all = TRUE)
# Set the NA (one-sided trips) to count of 0
res[is.na(res)] <- 0
```

```{r}
res$surplus <- res$Freq.y - res$Freq.x
res <- res[order(-res$surplus),]
res
```

```{r}
# Remove rows with no start station id, those are a result of the merge all=TRUE and should not exist
res <- res %>% filter(start_station_id != 'ABAN')
res
```

Above returns a list of 619887 combinations given every start-end combo exists as the inverse is valid, too except for Abandoned starts

So Freq.x is the count of trips from start to end
-- Freq.y is the count of trips from end to start
surplus is Freq.y - Freq.x which means that positive number indicates return trips are more and thus surplus of bikes
and negative number indicates return trips are less and thus the station is not breaking even

```{r}
res %>% filter(end_station_id == 'ABAN')
```

1363 rows indicate most start stations will have at least one abandoned bike

```{r}
res %>% filter(Freq.x == 0)
```

With Freq.x ==0 , there are 148,574  rows in which an end-to-start occurs, so the rows are pure surplus to the start_station_id

```{r}
# Want to identify the surplus (or not) by station for the month
station_surplus_count <- res %>%
  group_by(start_station_id) %>%
  summarize(surplus_sum=sum(surplus))
station_surplus_count
```
The above indicates 1656 rows, 2 less than the count of end_station_id from the initial dataset

```{r}
station_surplus_count <- station_surplus_count[order(-station_surplus_count$surplus_sum),]
#station_surplus_count %>%
#    filter(startsWith(as.character(start_station_id), 'S'))
colnames(station_surplus_count)[1] <- "station_id"
station_surplus_count
```
21 start with J
14 start with H

Above is result showing the surplus(+) or not surplus (-) of 1656 stations in the NYC district, this probably includes a few JC or HB stations

```{r}
sum(station_surplus_count$surplus_sum)
```

-8012 indicates 8012 abandons ... AND THAT IS A CORRECT NUMBER!!! phew!!!

```{r}
# Extract just the end station Id, lat, log
# because this had the higher count from the initial dataset, going with end_station_id
end_station_info <- citibike %>%
  select(end_station_id, end_lng, end_lat)

#end_station_info <- unique( end_station_info[ , 1:3 ] )
end_station_info <- end_station_info[!duplicated(end_station_info$end_station_id),]
end_station_info %>%
  filter(end_station_id == 'ABAN')
```

For awareness, the ABAN entry does have lng and lat
end_station_id
<chr>
end_lng
<dbl>
end_lat
<dbl>
ABAN	-73.99	40.67	

```{r}
# Now I want the coordinates of all those station_ids
station_surplus_count_coords <- merge(x = station_surplus_count, y = end_station_info, by.x = 'station_id', by.y='end_station_id', all.x = TRUE)

colnames(station_surplus_count_coords)[3] <- "lng"
colnames(station_surplus_count_coords)[4] <- "lat"
station_surplus_count_coords


station_surplus_count_coords <- station_surplus_count_coords %>%
  add_row(station_id = "ABAN", surplus_sum=1363, lng=-73.99, lat=40.67)

nrow(station_surplus_count_coords)
```
Above table accounts for all the station IDs with overal surplus or not surplus along with coordinates


```{r}
res
```

310,625 * 2 - 619,887 = 1363 (number of abandon rows ... this is correct)

```{r}
# This removed the duplicate rows by transposing the start and end station ids
res_temp <- res %>% select(start_station_id, end_station_id)


res2 <- res_temp[!duplicated(lapply(as.data.frame(t(res_temp), stringsAsFactors=FALSE), sort)),]
res2
```
So now there are 310625 edges in the graph, a little over half which makes sense

so now I need to add back on the surplus by merging ... actually I don't need to add the values, these are the edges and that should be sufficient.

```{r}
# Nodes
#station_surplus_count_coords
# station_id    surplus_sum   lng lat


# Edges
# res2
#start_station_id     end_station_id
```



```{r}
g_citi <- graph_from_data_frame(res2, directed=FALSE, vertices=station_surplus_count_coords)
```

Nodes
id lon lat name

Edges
from to weight category

```{r}
edges_for_plot <- res2 %>%
  inner_join(station_surplus_count_coords %>% select(station_id, lng, lat), by=c('start_station_id' = 'station_id')) %>%
  rename(x=lng, y=lat) %>%
  inner_join(station_surplus_count_coords %>% select(station_id, lng, lat), by=c('end_station_id' = 'station_id')) %>%
  rename(xend=lng, yend=lat)

assert_that(nrow(edges_for_plot) == nrow(res2))



```

```{r}
#nodes$weight = degree(g)
```

```{r}
maptheme <- theme(panel.grid = element_blank()) +
  theme(axis.text = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(axis.title = element_blank()) +
  theme(legend.position = "bottom") +
  theme(panel.grid = element_blank()) +
  theme(panel.background = element_rect(fill="#596673")) +
  theme(plot.margin = unit(c(0,0,0.5,0), 'cm'))
```

```{r}
country_shapes <- geom_polygon(aes(x=long, y=lat, group=group),
                               data=map_data('world'),
                               fill="#CECECE", color="#515151",
                               size=0.15)
mapcoords <- coord_fixed(xlim = c(-75, -73), ylim = c(40.5, 41))
```

```{r}
ggplot(station_surplus_count_coords) + country_shapes +
  geom_curve(aes(x = x, y = y, xend = xend, yend = yend),     # draw edges as arcs
             data = edges_for_plot, curvature = 0.33,
             alpha = 0.5) +
  scale_size_continuous(guide = FALSE, range = c(0.25, 2)) + # scale for edge widths
  geom_point(aes(x = lng, y = lat),           # draw nodes
             shape = 21, fill = 'white',
             color = 'black', stroke = 0.5) +
  scale_size_continuous(guide = "none", range = c(1, 6)) +    # scale for node size
  geom_text(aes(x = lng, y = lat, label = station_id),             # draw text labels
            hjust = 0, nudge_x = 1, nudge_y = 4,
            size = 3, color = "white", fontface = "bold") +
  mapcoords + maptheme
```
Well, there it is. Probably should be better, though.

# VALUE IN ASSESSING THE NETWORK GRAPH OF THE DATA, CLUSTER BY EDGES

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
