---
title: "Scratch Clean Station Metadata"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

Attempting with just my neighborhood, BK61 Crown Heights North

```{r libraries, echo=F, message=F, warning=F, include=F}
# Required packages
library(data.table)
library(tidyverse)
library(ggplot2)
library(skimr)
library(lubridate)
library(jsonlite)
library(MASS)
#library(assertthat)
#library(igraph)
#library(ggraph)
#library(ggmap)
#library(leaflet)
#library(rgdal)
#library(RColorBrewer)
```


```{r surplus.by.boro, echo=F, message=F, warning=F, include=T}
stations_with_elevation <- read.csv('stations_with_elevation.csv', row.names = 1, header= TRUE)

stations_with_boro_hood <- read.csv('stations_with_boro_and_hood.csv', row.names = 1, header= TRUE)

# Combine the elevation, borough, neighborhood, and September surplus
# First let's trim the DF for elevation
stations_with_elevation_trim <- stations_with_elevation %>%
  dplyr::select(short_name, station_id, elevation, elev_units)

stations_with_boro_hood_trim <- stations_with_boro_hood %>%
  dplyr::select(short_name, name, station_id, capacity, ntaname, boro_name, lon, lat)

stations_attrs_trim <- 
  merge(stations_with_elevation_trim, 
        stations_with_boro_hood_trim, 
        by.x=c('short_name'), 
        by.y=c('short_name'), all = TRUE)

stations_attrs_trim <- stations_attrs_trim %>%
  dplyr::select(-station_id.y)

colnames(stations_attrs_trim)[colnames(stations_attrs_trim) == 'station_id.x'] <- 'station_id'

stations_attrs_trim[c("boro_name")][is.na(stations_attrs_trim[c("boro_name")])] <- "New Jersey"

stations_attrs_trim <- 
  stations_attrs_trim %>% 
  mutate(ntaname = ifelse(startsWith(short_name, "JC"), "Jersey City", ntaname))

stations_attrs_trim <- 
  stations_attrs_trim %>% 
  mutate(ntaname = ifelse(startsWith(short_name, "HB"), "Hoboken", ntaname))

stations_attrs_trim <- stations_attrs_trim %>%
  rename(short.name=short_name, station.id=station_id, elev.units=elev_units, nta.name=ntaname, boro.name=boro_name)
  

head(stations_attrs_trim)
```

```{r}
stations_attrs_trim_bk <- stations_attrs_trim %>%
#stations_attrs_trim_bk61 <- stations_attrs_trim %>%
  #filter(nta.name == "Crown Heights North")
  filter(boro.name == "Brooklyn")
(stations_attrs_trim_bk)
#11 stations in this file
```

474 stations in brooklyn

```{r}
stations_with_bike_avail <- read.csv('bike_avail_by_station_and_time.csv', row.names = 1, header= TRUE, check.names = FALSE)
```

```{r}
# extract column names
colnames_bike_avail <- colnames(stations_with_bike_avail)
```

1802 columns, remove 1 (timestamp), so this means a total of 1801 station numbers

```{r}
stations_missing_from_metadata <- setdiff(colnames_bike_avail[-1],stations_attrs_trim$station.id)
```

104 are in the availability file but not in the station metadata file

```{r}
stations_missing_from_metadata
```

```{r}
# Let's try the clustering
# https://www.r-bloggers.com/2019/06/hierarchical-clustering-for-location-based-strategy-using-r-for-e-commerce/
library(geosphere)

# Distance matrix for docking stations
# Result is in meters
dist_mat <- distm(stations_attrs_trim_bk[9:10], stations_attrs_trim_bk[9:10], fun=distHaversine)
dist_mat <- as.data.frame(dist_mat)
```

```{r}
dist_mat[is.na(dist_mat)] <- 0

dMat <- as.dist(dist_mat)
```

```{r}
dMat[1:10]
```

```{r}
# Now for the clustering
hier_clust <- hclust(dMat, method = "complete")
# 400 meters is about a quarter of a mile (0.248548 miles)
stations_attrs_trim_bk$cluster <- cutree(hier_clust, h=400)
```

```{r}
head(stations_attrs_trim_bk)
```

213 clusters of 474 stations

1700 / 733 is 2.32, so ~2 docking stations per cluster

```{r}
# Predict with specific lat long
# https://stackoverflow.com/questions/27321856/closest-point-to-a-path
library(RANN)
#40.67355733701634, -73.96085900000001
# Lon, Lat of location
location <- data.frame(matrix(nrow = 1,data = c(-73.960859, 40.67355)))
closest_station <- nn2(stations_attrs_trim_bk61[, 9:10], query=location, k=1)
inp_cluster <- stations_attrs_trim_bk61[closest_station$nn.idx,]$cluster
inp_cluster
```

```{r}
# Lat/long of 790 Classon 40.652256, -73.956582
```

```{r}
# calculating a nearest neighbor
#distance <- sqrt(rowSums((df-do.call(rbind,replicate(nrow(df),p,simplify = FALSE)))**2))
#nearest <- df[which.min(distance),]
# Instead, follow the steps here:
# https://stackoverflow.com/questions/21977720/r-finding-closest-neighboring-point-and-number-of-neighbors-within-a-given-rad
```

```{r}
stations_with_bike_avail
```

```{r}
stations_with_bike_avail_long <- stations_with_bike_avail %>%
  pivot_longer(!timestamp, names_to = "station.id", values_to = "bikes.avail.count")
```
Pivot longer results in 2418743 rows

```{r}
station_to_cluster <- stations_attrs_trim_bk %>%
  dplyr::select(station.id, cluster)

(station_to_cluster)
```

```{r}
# station_to_cluster contains the cluster number for each station
# So now merge with stations_with_bike_avail_long
```

```{r}

```

```{r}
# Merge long df with counts with station.id and cluster to label clusters properly
stations_bike_avail_by_time <- 
  merge(stations_with_bike_avail_long, 
        station_to_cluster, 
        by.x=c('station.id'), 
        by.y=c('station.id'), all.x = TRUE)
```

```{r}
stations_bike_avail_by_time_no_na <- stations_bike_avail_by_time %>%
  filter(!is.na(cluster))
```
139,672 rows without a cluster because there are more station.id values from API compared to station information json which contained the lat and long

```{r}
stations_bike_avail_by_time_no_na
```

```{r}
# group by cluster ID for each timestamp
stations_bike_avail_by_time_no_na_group <- aggregate(bikes.avail.count ~ timestamp + cluster, data = stations_bike_avail_by_time_no_na, FUN = sum, na.rm = TRUE)
```

```{r}
class(stations_bike_avail_by_time_no_na_group$timestamp)
```

```{r}
stations_bike_avail_by_time_no_na_group <- stations_bike_avail_by_time_no_na_group[order(stations_bike_avail_by_time_no_na_group$timestamp),]
```

```{r}
stations_bike_avail_by_time_no_na_group
```

```{r}
# mutate to convert timestamp into day of the week, etc.
stations_bike_avail_by_time_no_na_group <- stations_bike_avail_by_time_no_na_group %>%
  mutate(time=as.ITime(ymd_hms(timestamp)),
         weekday = lubridate::wday(ymd_hms(timestamp),label=TRUE,abbr=TRUE))
```

```{r}
stations_bike_avail_by_time_no_na_group
```

```{r}
skim(stations_bike_avail_by_time_no_na_group)
```

```{r}
# https://stats.oarc.ucla.edu/r/dae/negative-binomial-regression/
```

```{r}
summary(m1 <- glm.nb(bikes.avail.count ~ cluster + time + weekday, data = stations_bike_avail_by_time_no_na_group))
```

```{r}
m2 <- update(m1, . ~ . - weekday)
anova(m1, m2)
```

```{r}
m3 <- glm(bikes.avail.count ~ cluster + time + weekday, family = "poisson", data = stations_bike_avail_by_time_no_na_group)
pchisq(2 * (logLik(m1) - logLik(m3)), df = 1, lower.tail = FALSE)
```

```{r}
stations_bike_avail_by_time_no_na_group$cluster <- as.factor(stations_bike_avail_by_time_no_na_group$cluster)
stations_bike_avail_by_time_no_na_group$time <- as.factor(stations_bike_avail_by_time_no_na_group$time)
stations_bike_avail_by_time_no_na_group$weekday <- as.factor(stations_bike_avail_by_time_no_na_group$weekday)

stations_bike_avail_by_time_no_na_group$time <- factor(stations_bike_avail_by_time_no_na_group$time, 
                       levels = c("00:00:00", "00:15:00", "00:30:00", "00:45:00",
                                  "01:00:00", "01:15:00", "01:30:00", "01:45:00",
                                  "02:00:00", "02:15:00", "02:30:00", "02:45:00",
                                  "03:00:00", "03:15:00", "03:30:00", "03:45:00",
                                  "04:00:00", "04:15:00", "04:30:00", "04:45:00",
                                  "05:00:00", "05:15:00", "05:30:00", "05:45:00",
                                  "06:00:00", "06:15:00", "06:30:00", "06:45:00",
                                  "07:00:00", "07:15:00", "07:30:00", "07:45:00",
                                  "08:00:00", "08:15:00", "08:30:00", "08:45:00",
                                  "09:00:00", "09:15:00", "09:30:00", "09:45:00",
                                  "10:00:00", "10:15:00", "10:30:00", "10:45:00",
                                  "11:00:00", "11:15:00", "11:30:00", "11:45:00",
                                  "12:00:00", "12:15:00", "12:30:00", "12:45:00",
                                  "13:00:00", "13:15:00", "13:30:00", "13:45:00",
                                  "14:00:00", "14:15:00", "14:30:00", "14:45:00",
                                  "15:00:00", "15:15:00", "15:30:00", "15:45:00",
                                  "16:00:00", "16:15:00", "16:30:00", "16:45:00",
                                  "17:00:00", "17:15:00", "17:30:00", "17:45:00",
                                  "18:00:00", "18:15:00", "18:30:00", "18:45:00",
                                  "19:00:00", "19:15:00", "19:30:00", "19:45:00",
                                  "20:00:00", "20:15:00", "20:30:00", "20:45:00",
                                  "21:00:00", "21:15:00", "21:30:00", "21:45:00",
                                  "22:00:00", "22:15:00", "22:30:00", "22:45:00",
                                  "23:00:00", "23:15:00", "23:30:00", "23:45:00"))

stations_bike_avail_by_time_no_na_group
```

```{r}
# following these steps
# https://www.r-bloggers.com/2020/01/count-data-models/
```

```{r}
tab <- table(stations_bike_avail_by_time_no_na_group$bikes.avail.count)
tab
# Below shows the range of values for `bikes.avail.count` with 35611 at zero and max of 438 with one occurrence
```

```{r}
stations_bike_avail_by_time_no_timestamp <- subset(stations_bike_avail_by_time_no_na_group, select=-c(timestamp))
```

```{r}
stations_bike_avail_by_time_no_timestamp
```

```{r}
stations_bike_avail_by_time_no_timestamp_1 <- modify_if(stations_bike_avail_by_time_no_timestamp, is.factor, as.integer)
```

```{r}
library(corrr)
M <- correlate(stations_bike_avail_by_time_no_timestamp_1, method="spearman")
rplot(shave(M), colours=c("red", "white", "blue" ))+
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# Start with data partition here.
set.seed(8675309)
index <- sample(2, nrow(stations_bike_avail_by_time_no_timestamp), replace = TRUE, p=c(0.8, 0.2))
train <- stations_bike_avail_by_time_no_timestamp[index==1,]
test <- stations_bike_avail_by_time_no_timestamp[index==2,]
```

```{r}
# Sample Variance is 313
var(stations_bike_avail_by_time_no_timestamp$bikes.avail.count)
n <- length(stations_bike_avail_by_time_no_timestamp)
# Population variance
var(stations_bike_avail_by_time_no_timestamp$bikes.avail.count)*(n-1)/n
# Mean is 13
mean(stations_bike_avail_by_time_no_timestamp$bikes.avail.count)

# Overdispersion exists
```

```{r}
train
```


Poisson Model: Generalized linear model to model count data, assumption that variance is equal to the mean

```{r}
mod1 <- glm(bikes.avail.count ~ cluster + time + weekday, data = train, family ="poisson")
summary(mod1)
```

```{r}
library(broom)
glance(mod1)
```


```{r}
library(AER)
dispersiontest(mod1)
# This indicates dispersion
```

Quasi-Poisson Model: Generalized linear model to model count data, for overdispersed count variable
Variance is a linear function of the mean

```{r}
mod2 <- glm(bikes.avail.count ~ cluster + time + weekday, data = train, family ="quasipoisson")
summary(mod2)
```

```{r}
pred <- predict.glm(mod1, newdata=test[test$bikes.avail.count != 0,],type = "response")
plot(test$bikes.avail.count[test$bikes.avail.count!=0],type = "b",col="red")
lines(round(pred),col="blue")
```



```{r}
library(Metrics)
pred <- predict.glm(mod1, newdata = test, type = "response")
rmsemodelp <- ModelMetrics::rmse(test$bikes.avail.count,round(pred))
maemodelp <- mae(test$bikes.avail.count,round(pred))
rmsemodelp
maemodelp
```


```{r}
predq <- predict.glm(mod2, newdata=test[test$bikes.avail.count != 0,],type = "response")
plot(test$bikes.avail.count[test$bikes.avail.count!=0],type = "b",col="red")
lines(round(predq),col="blue")
```

```{r}
predq <- predict.glm(mod2,newdata=test, type = "response")
rmsemodelqp <- ModelMetrics::rmse(test$bikes.avail.count,round(predq))
maemodelqp <- mae(test$bikes.avail.count,round(predq))
rmsemodelqp
maemodelqp
```

Negative Binomial: Variance greater than the mean
Variance is a quadratic function of the mean

```{r}
mod3 <- glm.nb(bikes.avail.count ~ cluster + time + weekday, data=train)
summary(mod3)
```

```{r}
prednb <- predict.glm(mod3,newdata=test[test$bikes.avail.count!=0,],type = "response")
plot(test$bikes.avail.count[test$bikes.avail.count!=0],type = "b",col="red")
lines(round(prednb),col="blue")
```

```{r}
prednb <- predict.glm(mod3,newdata=test,type = "response")
rmsemodelnb <- ModelMetrics::rmse(test$bikes.avail.count,round(prednb))
maemodelnb <- mae(test$bikes.avail.count,round(prednb))
rmsemodelnb
maemodelnb
```

Hurdle Poisson

Two parts: Probability of attaining value of 0, and second part models probability of a non-zero value

```{r}
library(pscl)
modelhp <- hurdle(bikes.avail.count ~ cluster + time + weekday, data=train, dist = "poisson")
summary(modelhp)
```

```{r}
predhp <- predict(modelhp, newdata=test[test$bikes.avail.count!=0,],type = "response")
plot(test$bikes.avail.count[test$bikes.avail.count!=0],type = "b",col="red")
lines(round(predhp),col="blue")
```

```{r}
predhp <- predict(modelhp, newdata=test, type = "response")
rmsemodelhp<-ModelMetrics::rmse(test$bikes.avail.count,round(predhp))
maemodelhp<-mae(test$bikes.avail.count,round(predhp))
rmsemodelhp
maemodelhp
```

Hurdle Negative Binomial

```{r}
modelhnb <- hurdle(bikes.avail.count ~ cluster + time + weekday, data=train, dist = "negbin")
summary(modelhnb)
```

```{r}
predhnb <- predict(modelhnb, newdata=test[test$bikes.avail.count!=0,],type = "response")
plot(test$bikes.avail.count[test$bikes.avail.count!=0],type = "b",col="red")
lines(round(predhnb),col="blue")
```

```{r}
predhnb<- predict(modelhnb, newdata=test,type = "response")
rmsemodelhnb<-ModelMetrics::rmse(test$bikes.avail.count,round(predhnb))
maemodelhnb<-mae(test$bikes.avail.count,round(predhnb))
rmsemodelhnb
maemodelhnb
```

Zero Inflated Model: Distribution that allows for frequent zero-values observations

```{r}
modelzp <- zeroinfl(bikes.avail.count ~ cluster + time + weekday, data=train, dist = "poisson")
summary(modelzp)
```

```{r}
predzp<- predict(modelzp,newdata=test,type = "response")
rmsemodelzp<-ModelMetrics::rmse(test$bikes.avail.count,round(predzp))
maemodelzp<-mae(test$bikes.avail.count,round(predzp))
rmsemodelzp
maemodelzp
```

```{r}
modelznb <- zeroinfl(bikes.avail.count ~ cluster + time + weekday, data=train,dist = "negbin")
summary(modelznb)
```

```{r}
predznb<- predict(modelznb,newdata=test,type = "response")
rmsemodelznb<-ModelMetrics::rmse(test$bikes.avail.count,round(predznb))
maemodelznb<-mae(test$bikes.avail.count,round(predznb))
rmsemodelznb
maemodelznb
```


```{r}
rmse<-c(rmsemodelp,rmsemodelqp,rmsemodelnb,rmsemodelhp,rmsemodelhnb,
           rmsemodelzp,rmsemodelznb)
mae<-c(maemodelp,maemodelqp,maemodelnb,maemodelhp,maemodelhnb,
           maemodelzp,maemodelznb)
models<-c("pois","q_pois","nb","h_pois","h_nb","zer_pois","zer_nb")
data.frame(models,rmse,mae)%>% 
  arrange(rmse)
```

```{r}
# inp_cluster
bike_avail_query <- data.frame(matrix(nrow = 1,data = c(inp_cluster, 0, "07:30:00", "Mon")))
colnames(bike_avail_query) <- c("cluster", "bikes.avail.count", "time", "weekday")

single_pred <- predict.glm(mod1, newdata = bike_avail_query, type = "response")
as.integer(floor(single_pred))

```

```{r}
predict_num_bikes_avail <- function(longitude, latitude, time, day) {
  #longitude: -73.960859
  #latitude: 40.67355
  #time: "07:30:00"
  #day: "Mon"
  
  # Convert lon/lat to cluster
  location <- data.frame(matrix(nrow = 1,data = c(longitude, latitude)))
  closest_station <- nn2(stations_attrs_trim_bk[, 9:10], query=location, k=1)
  inp_cluster <- stations_attrs_trim_bk[closest_station$nn.idx,]$cluster

  print(inp_cluster)
  
  # Create input for prediction
  bike_avail_query <- data.frame(matrix(nrow = 1,data = c(inp_cluster, 0, time, day)))
  colnames(bike_avail_query) <- c("cluster", "bikes.avail.count", "time", "weekday")

  # Predict using Model: mod1
  num_bikes_avail <- predict.glm(mod1, newdata = bike_avail_query, type = "response")
  # Take floor of prediction to ensure whole number
  num_bikes_avail <- as.integer(floor(num_bikes_avail))
  
  return(num_bikes_avail)
}
```

```{r}
predict_num_bikes_avail(-73.960859, 40.67355, "08:30:00", "Sat")
```

# Attempting by hour instead of 15-minute interval

```{r}
citibike_by_end.interval_1H <- stations_bike_avail_by_time_no_na_group %>%
  mutate(interval=lubridate::floor_date(as.POSIXct(timestamp), "1 hour")) %>%
  group_by(cluster, interval) %>%
  summarize(sum(bikes.avail.count))

```

```{r}
stations_bike_avail_by_time_no_na_group
```

```{r}

```

```{r}
stations_bike_avail_by_time_1H <- stations_bike_avail_by_time_no_timestamp %>%
  filter(endsWith(as.character(time), "00:00"))
```

```{r}
stations_bike_avail_by_time_1H
```

```{r}
# Start with data partition here.
set.seed(8675309)
index <- sample(2, nrow(stations_bike_avail_by_time_1H), replace = TRUE, p=c(0.8, 0.2))
train_1H <- stations_bike_avail_by_time_1H[index==1,]
test_1H <- stations_bike_avail_by_time_1H[index==2,]
```

```{r}
# Sample Variance is 313
var(stations_bike_avail_by_time_1H$bikes.avail.count)
n <- length(stations_bike_avail_by_time_1H)
# Population variance
var(stations_bike_avail_by_time_1H$bikes.avail.count)*(n-1)/n
# Mean is 13
mean(stations_bike_avail_by_time_1H$bikes.avail.count)

# Overdispersion exists
```

```{r}

```


Poisson Model: Generalized linear model to model count data, assumption that variance is equal to the mean

```{r}
mod1 <- glm(bikes.avail.count ~ cluster + time + weekday, data = train_1H, family ="poisson")
summary(mod1)
```

```{r}
glance(mod1)
```


```{r}
dispersiontest(mod1)
# This indicates dispersion
```

Quasi-Poisson Model: Generalized linear model to model count data, for overdispersed count variable
Variance is a linear function of the mean

```{r}
mod2 <- glm(bikes.avail.count ~ cluster + time + weekday, data = train_1H, family ="quasipoisson")
summary(mod2)
```

```{r}
pred <- predict.glm(mod1, newdata=test_1H[test_1H$bikes.avail.count != 0,],type = "response")
plot(test_1H$bikes.avail.count[test_1H$bikes.avail.count!=0],type = "b",col="red")
lines(round(pred),col="blue")
```



```{r}
library(Metrics)
pred <- predict.glm(mod1, newdata = test_1H, type = "response")
rmsemodelp <- ModelMetrics::rmse(test_1H$bikes.avail.count,round(pred))
maemodelp <- mae(test_1H$bikes.avail.count,round(pred))
rmsemodelp
maemodelp
```


```{r}
predq <- predict.glm(mod2, newdata=test_1H[test_1H$bikes.avail.count != 0,],type = "response")
plot(test_1H$bikes.avail.count[test_1H$bikes.avail.count!=0],type = "b",col="red")
lines(round(predq),col="blue")
```

```{r}
predq <- predict.glm(mod2,newdata=test_1H, type = "response")
rmsemodelqp <- ModelMetrics::rmse(test_1H$bikes.avail.count,round(predq))
maemodelqp <- mae(test_1H$bikes.avail.count,round(predq))
rmsemodelqp
maemodelqp
```

Negative Binomial: Variance greater than the mean
Variance is a quadratic function of the mean

```{r}
mod3 <- glm.nb(bikes.avail.count ~ cluster + time + weekday, data=train_1H)
summary(mod3)
```

```{r}
prednb <- predict.glm(mod3,newdata=test_1H[test_1H$bikes.avail.count!=0,],type = "response")
plot(test_1H$bikes.avail.count[test_1H$bikes.avail.count!=0],type = "b",col="red")
lines(round(prednb),col="blue")
```

```{r}
prednb <- predict.glm(mod3,newdata=test_1H,type = "response")
rmsemodelnb <- ModelMetrics::rmse(test_1H$bikes.avail.count,round(prednb))
maemodelnb <- mae(test_1H$bikes.avail.count,round(prednb))
rmsemodelnb
maemodelnb
```

```{r}
modelhp <- hurdle(bikes.avail.count ~ cluster + time + weekday, data=train_1H, dist = "poisson")
summary(modelhp)
```

```{r}
predhp <- predict(modelhp, newdata=test_1H[test_1H$bikes.avail.count!=0,],type = "response")
plot(test_1H$bikes.avail.count[test_1H$bikes.avail.count!=0],type = "b",col="red")
lines(round(predhp),col="blue")
```

```{r}
predhp <- predict(modelhp, newdata=test_1H, type = "response")
rmsemodelhp<-ModelMetrics::rmse(test_1H$bikes.avail.count,round(predhp))
maemodelhp<-mae(test_1H$bikes.avail.count,round(predhp))
rmsemodelhp
maemodelhp
```

Hurdle Negative Binomial

```{r}
modelhnb <- hurdle(bikes.avail.count ~ cluster + time + weekday, data=train_1H, dist = "negbin")
summary(modelhnb)
```

```{r}
predhnb <- predict(modelhnb, newdata=test_1H[test_1H$bikes.avail.count!=0,],type = "response")
plot(test_1H$bikes.avail.count[test_1H$bikes.avail.count!=0],type = "b",col="red")
lines(round(predhnb),col="blue")
```

```{r}
predhnb<- predict(modelhnb, newdata=test_1H,type = "response")
rmsemodelhnb<-ModelMetrics::rmse(test_1H$bikes.avail.count,round(predhnb))
maemodelhnb<-mae(test_1H$bikes.avail.count,round(predhnb))
rmsemodelhnb
maemodelhnb
```

Zero Inflated Model: Distribution that allows for frequent zero-values observations

```{r}
modelzp <- zeroinfl(bikes.avail.count ~ cluster + time + weekday, data=train_1H, dist = "poisson")
summary(modelzp)
```

```{r}
predzp<- predict(modelzp,newdata=test_1H,type = "response")
rmsemodelzp<-ModelMetrics::rmse(test_1H$bikes.avail.count,round(predzp))
maemodelzp<-mae(test_1H$bikes.avail.count,round(predzp))
rmsemodelzp
maemodelzp
```

```{r}
modelznb <- zeroinfl(bikes.avail.count ~ cluster + time + weekday, data=train_1H,dist = "negbin")
summary(modelznb)
```

```{r}
predznb<- predict(modelznb,newdata=test_1H,type = "response")
rmsemodelznb<-ModelMetrics::rmse(test_1H$bikes.avail.count,round(predznb))
maemodelznb<-mae(test_1H$bikes.avail.count,round(predznb))
rmsemodelznb
maemodelznb
```


```{r}
rmse<-c(rmsemodelp,rmsemodelqp,rmsemodelnb,rmsemodelhp,rmsemodelhnb,
           rmsemodelzp,rmsemodelznb)
mae<-c(maemodelp,maemodelqp,maemodelnb,maemodelhp,maemodelhnb,
           maemodelzp,maemodelznb)
models<-c("pois","q_pois","nb","h_pois","h_nb","zer_pois","zer_nb")
data.frame(models,rmse,mae)%>% 
  arrange(rmse)
```

# I want to plot the stations and color code by cluster

stations_attrs_trim_bk

```{r}
library(pals)
# Create color factor object
cluster_color <- colorFactor(palette=polychrome(), domain=stations_attrs_trim_bk$cluster)

# Using basemaps for NYC
m <- leaflet(data=stations_attrs_trim_bk) %>% 
  addTiles() %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(cluster), 
    label=~as.character(cluster), 
    radius=.5,
    color = ~cluster_color(cluster)
  )


# Display map
m
```

```{r}
stations_attrs_trim_bk
```

```{r}
polychrome
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r eval=F}
tab_df <- as.data.frame(tab)

# Distrubtion of count results
p<-ggplot(data=tab_df, aes(x=Var1, y=Freq)) +
  geom_bar(stat="identity")
p
```

```{r}

```

