---
title: "Matrix of Start Stop by 15 minute intervals"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

# GOAL IS TO CREATE A MATRIX OF THE STARTS/ENDS FOR 15 MINUTE INTERVALS FOR EACH STATION

```{r}
# Required packages
library(tidyverse)
#library(ggplot2)
#library(skimr)
#library(tidyjson)
#library(geosphere)
library(lubridate)
# library(ggmap)
#library(forcats)
#library(scales)
```

```{r}
# Create additional columns of pertinence
# weekday, day of the month, trip duration in minutes, start hour
citibike_202208 <- fread("data/202208-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
citibike_202209 <- fread("data/202209-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
citibike_202210 <- fread("data/202210-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
#citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%

citibike <- bind_rows(citibike_202208, citibike_202209, citibike_202210) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = lubridate::wday(ymd_hms(started_at), label=TRUE, abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual)) %>%
  rename(ride.id=ride_id, rideable.type=rideable_type, started.at=started_at,
         ended.at=ended_at, start.station.name=start_station_name, start.station.id=start_station_id,
         end.station.name=end_station_name, end.station.id=end_station_id, start.lat=start_lat,
         start.lng=start_lng, end.lat=end_lat, end.lng=end_lng, member.casual=member_casual)

# Figure out abandoned and label them
citibike$end.station.name[citibike$end.station.name == ''] <- "Abandoned"
citibike$end.station.id[citibike$end.station.id  == ''] <- "ABAN"

# Output DF if needed
head(citibike)
```



```{r}
nrow(citibike)
```


```{r}
citibike_wed <- citibike %>%
  filter(weekday == "Wed" & start.station.id != "")
```

```{r}
nrow(citibike_wed)
```

```{r}
# And now for grouping by 15 minutes

citibike_by_start.interval <- citibike_wed %>%
  mutate(start.interval=lubridate::floor_date(started.at, "1 hour")) %>%
  group_by(start.station.id, start.interval) %>%
  summarize(start.count=n())

citibike_by_start.interval
```

```{r}
citibike_by_end.interval <- citibike_wed %>%
  mutate(end.interval=lubridate::floor_date(ended.at, "1 hour")) %>%
  group_by(end.station.id, end.interval) %>%
  summarize(count=n())

citibike_by_end.interval
```

```{r}
# Now I Need to merge by station_id and time interval

citibike_by_interval <- 
  merge(citibike_by_start.interval, 
        citibike_by_end.interval, 
        by.x=c('start.station.id','start.interval'), 
        by.y=c('end.station.id','end.interval'), all = TRUE)
```



```{r}
citibike_by_interval[is.na(citibike_by_interval)] <- 0
citibike_by_interval
```

```{r}
# Rename columns
colnames(citibike_by_interval)[1] <- "station_id"
colnames(citibike_by_interval)[2] <- "interval_start.time"
colnames(citibike_by_interval)[3] <- "depart.count"
colnames(citibike_by_interval)[4] <- "arrive.count"

citibike_by_interval
```

```{r}
# Calculate surplus
citibike_by_interval$surplus <- citibike_by_interval$arrive.count - citibike_by_interval$depart.count

citibike_by_interval
```

```{r}
citibike_by_interval_surplus <- citibike_by_interval %>%
  dplyr::select(station_id, interval_start.time, surplus)

citibike_by_interval_surplus
```

```{r}
mydateseq<-seq(as.POSIXct("2022-08-01"), by="1 hour", length.out = 2208)  

# Output the data sequence to confirm the result are 15 minute intervals which start on the hour
tail(mydateseq)
```

```{r}
citibike_by_interval_surplus_wide <- citibike_by_interval_surplus %>%
  pivot_wider(names_from = station_id, values_from = surplus)

citibike_by_interval_surplus_wide[is.na(citibike_by_interval_surplus_wide)] <- 0


# For ordering
citibike_by_interval_surplus_wide <- citibike_by_interval_surplus_wide[order(citibike_by_interval_surplus_wide$interval_start.time),]
citibike_by_interval_surplus_wide
```
# Nice the above matrix is by time and then for all the columns are stations with the plus/minus for each one hour interval

# 8P on 11/9/22, now I want to do is group by time with mean, regardless of day

```{r}
citibike_by_interval_surplus_wide$interval.start.ch <- as.character(citibike_by_interval_surplus_wide$interval_start.time)

head(citibike_by_interval_surplus_wide[,1685:1689])
```

```{r}
citibike_by_interval_surplus_wide$interval.start.ch <- sub(".*? ", "", citibike_by_interval_surplus_wide$interval.start.ch)
```

```{r}
citibike_by_interval_surplus_wide <- citibike_by_interval_surplus_wide %>%
  dplyr::select(-interval_start.time)
```


```{r}
citibikestation_by_hour <- citibike_by_interval_surplus_wide %>%
  pivot_longer(!interval.start.ch, names_to = "station.id", values_to = "count")
```

```{r}
citibikestation_by_hour %>%
  filter(count > 0)
```


```{r}
citibikestation_avg_by_hour <- citibikestation_by_hour %>%
  group_by(interval.start.ch, station.id) %>%
  summarize(avg=mean(count))

citibikestation_avg_by_hour
```

```{r}
citibikestation_avg_by_hour %>%
  filter(avg == 0)
```

The above number indicates that for one station at one hour, the average surplus/shortage across the entire 3 months of data on Wednesdays is avg column.

```{r}
stations_with_boro_hood <- read.csv('stations_with_boro_and_hood.csv', row.names = 1, header= TRUE)
```

```{r}
stations_with_boro_hood
```

```{r}
citibikestation_avg_by_hour_lat_long <- 
  merge(citibikestation_avg_by_hour, 
        stations_with_boro_hood, 
        by.x=c('station.id'), 
        by.y=c('short_name'), all.x = TRUE)
```


```{r}
citibikestation_avg_by_hour_lat_long <- citibikestation_avg_by_hour_lat_long %>%
  dplyr::select(interval.start.ch, station.id, avg, lon, lat)
```

```{r}
citibikestation_avg_by_hour_lat_long
```

```{r}
# Removed the NJ stations just to keep focus on NYC stations
citibikestation_avg_by_hour_lat_long_5A <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "05:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
citibikestation_avg_by_hour_lat_long_8A <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "08:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
citibikestation_avg_by_hour_lat_long_11A <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "11:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
```

```{r}
citibikestation_avg_by_hour_lat_long_2P <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "14:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
citibikestation_avg_by_hour_lat_long_5P <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "17:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
citibikestation_avg_by_hour_lat_long_8P <- citibikestation_avg_by_hour_lat_long %>%
  filter(interval.start.ch == "20:00:00" & !startsWith(station.id, "JC") & !startsWith(station.id, "HB"))
```

```{r}
threshold_p <- 0.5
threshold_n <- threshold_p * -1
```

```{r}
# Using basemaps for NYC
m_5A <- leaflet(data=citibikestation_avg_by_hour_lat_long_5A) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )

# Display map
m_5A
```

```{r}
# Using basemaps for NYC
m_8A <- leaflet(data=citibikestation_avg_by_hour_lat_long_8A) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )


# Display map
m_8A
```

```{r}
# Using basemaps for NYC
m_11A <- leaflet(data=citibikestation_avg_by_hour_lat_long_11A) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )


# Display map
m_11A
```

```{r}
# Using basemaps for NYC
m_2P <- leaflet(data=citibikestation_avg_by_hour_lat_long_2P) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )


# Display map
m_2P
```


```{r}
# Using basemaps for NYC
m_5P <- leaflet(data=citibikestation_avg_by_hour_lat_long_5P) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )


# Display map
m_5P
```

```{r}
# Using basemaps for NYC
m_8P <- leaflet(data=citibikestation_avg_by_hour_lat_long_8P) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~lon, ~lat, 
    popup=~as.character(station.id), 
    label=~as.character(station.id), 
    radius=.5,
    color = ~ifelse(avg > threshold_p, 'blue', 
                    ifelse(avg < threshold_n, 'red', 'green'))
  )


# Display map
m_8P
```

```{r}
library(mapview)
#Output leaflet maps to JPG
mapshot(m_5A, file = "cb_stations_surplus_5A.jpg")
mapshot(m_8A, file = "cb_stations_surplus_8A.jpg")
mapshot(m_11A, file = "cb_stations_surplus_11A.jpg")
mapshot(m_2P, file = "cb_stations_surplus_2P.jpg")
mapshot(m_5P, file = "cb_stations_surplus_5P.jpg")
mapshot(m_8P, file = "cb_stations_surplus_8P.jpg")
```

# IGNORE EVERYTHING BELOW ... 3:32P ON NOV 20, 2022

```{r}

```

```{r}

```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```


# Next is to count the running total

```{r}
citibike_by_interval_surplus_wide
```

```{r}
citibike_by_interval_surplus_wide_cs <- cumsum(citibike_by_interval_surplus_wide[2:1659])
citibike_by_interval_surplus_wide_cs
```

```{r}
time_index <- citibike_by_interval_surplus_wide$interval_start.time
time_index
```

```{r}
citibike_by_interval_surplus_wide
```
```{r}
# Write the above to CSV
#write.csv(citibike_by_interval_surplus_wide, file = 'stations_sept22_interval_surplus.csv')
```

```{r}

citibike_by_interval_surplus_wide_cs <- cbind(time_index, citibike_by_interval_surplus_wide_cs)
citibike_by_interval_surplus_wide_cs
```

Alright, so `citibike_by_interval_surplus_wide_cs` is the running total of eash station

```{r}
tail(citibike_by_interval_surplus_wide_cs)
```
# So many of the stations are negative, which can't be possible. Zero is absolutely the lowest number of bikes

```{r}
ff = function(x)
{
    cs = cumsum(x)
    if (cs < 0) {
      cs = 0
    } else if (cs > 20) {
      cs = 20
    }
}
ff(citibike_by_interval_surplus_wide[2])
```

```{r}
cumsum.bounded <- function(x, lower.bound = 0, upper.bound = 500) {
    bsum <- function(x, y) min(upper.bound, max(lower.bound, x+y))
    if (length(x) > 1) Reduce(bsum, x, acc = TRUE) else x
}
```

```{r}
cumsum.bounded(citibike_by_interval_surplus_wide[2:1659], 0, 20)
```

```{r}

```

```{r}
library(dm)
```

```{r}
cumsum_bounded(citibike_by_interval_surplus_wide[2:1659], 0, 20)
```

Unfortunately I wasn't able to get the cumsum bounded ... yet, need to check the .cpp suggestion

https://stackoverflow.com/questions/21262424/bounded-cumulative-sum

https://stat.ethz.ch/pipermail/r-help/2010-November/261244.html

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
