---
title: "DATA 698 Project v1"
subtitle: "Predicting Citi Bike Availability in NYC"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "default"
    fonttheme: "structurebold"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Problem and Objective (1)

Introduce yourselves and describe your problem.
Explain your objectives, challenges of your work, proposed methodologies, and the assumptions you made while conducting modeling and/or analysis.
Provide an overview of your approach and/or conceptual model (please do not present your code directly). 
Describe the results you obtain and summarize the current achievements and possibility of future works.

v2 is started at 1:30P on Sat, Nov 19

## Previous Work (+1)

- bullet 1
- bullet 2
- bullet 3
- bullet 4

## Challenges and Assumptions (2)

- Too much data
  - Over 3.5 million trips in month of Sept. 2022
- Rebalancing identification
- User friendly approach
  - Inputs to output
  
## Data Collection (3)

- Creating timetable of surplus/shortage

## Data Analysis (4)


## EDA 1 (5)

```{r libraries, echo=F, message=F, warning=F, include=F}
# Required packages
library(tidyverse)
library(ggplot2)
library(skimr)
library(lubridate)
library(fpp3)
library(assertthat)
library(igraph)
library(ggraph)
library(ggmap)
library(leaflet)
library(rgdal)
library(RColorBrewer)
library(jpeg)
library(data.table)
```

```{r read.trip.data, echo=F, message=F, warning=F, include=F, eval=T}
# Create additional columns of pertinence
# weekday, day of the month, trip duration in minutes, start hour
citibike_202208 <- fread("data/202208-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
citibike_202209 <- fread("data/202209-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
citibike_202210 <- fread("data/202210-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE)
#citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%

citibike <- bind_rows(citibike_202208, citibike_202209, citibike_202210) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = lubridate::wday(ymd_hms(started_at), label=TRUE, abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual)) %>%
  rename(ride.id=ride_id, rideable.type=rideable_type, started.at=started_at,
         ended.at=ended_at, start.station.name=start_station_name, start.station.id=start_station_id,
         end.station.name=end_station_name, end.station.id=end_station_id, start.lat=start_lat,
         start.lng=start_lng, end.lat=end_lat, end.lng=end_lng, member.casual=member_casual)

# Figure out abandoned and label them
citibike$end.station.name[citibike$end.station.name == ''] <- "Abandoned"
citibike$end.station.id[citibike$end.station.id  == ''] <- "ABAN"

# Output DF if needed
head(citibike)
```

```{r}
head(citibike)
```

10,210,102 trips for Aug-Oct (3 months of 2022)

```{r trip.by.hour.time.series, echo=T, message=F, warning=F, include=T}

citibike$started.at.ts <- as_datetime(as.character(citibike$started.at))

citibike_by_hour <- citibike %>%
  mutate(hour=lubridate::floor_date(started.at.ts, "1 hour")) %>%
  group_by(hour, member.casual) %>%
  summarize(cnt=n())
#citibike_by_hour

citibike_by_hour_ts <- citibike_by_hour %>%
  as_tsibble(index=hour, key=c(member.casual))
#citibike_by_hour_ts

autoplot(citibike_by_hour_ts, cnt) +
  labs(title = "Ride Trips by Hour - Sept. 2022",
       subtitle = "Citi Bike NYC",
       y = "Ride Trips Counts")
```

## EDA 2 (6)

```{r trip.by.hour.seasonal, echo=T, message=F, warning=F, include=T}
citibike_by_hour_ts %>%
  gg_season(cnt, period = "week") +
  labs(y = "Ride Trips",
       title = "Seasonal plot: Weekly Trip Counts for Sept. 2022")
```

## EDA 3 (7)

```{r display.surplus.by.station, echo=T, message=F, warning=F, include=T}
img <- readJPEG("stations_surplus_sum.jpg")
plot(1:10,ty="n", axes = 0, xlab='', ylab='', main='Overall Monthly Surplus by Docking Station')
rasterImage(img,-1,-1,12,12)
```

## EDA 4 (8)

Elevation or Network Graph


## focus on Brooklyn Areas (9)

```{r}

```

## Proposed Methodologies (10)

- Inputs
  - Latitude and longitude
  - Day of the Week
  - Time of Day
- Citi Bike offers live map of availability
- Lyft provides real-time availability

- Time series model
- Poisson distribution and Negative Binomial given the over-dispersion

## Overview of Approach (11)
 
- Data from API call every 15 minutes for two weeks
  - Citi bike availability at each station
  - Three weeks is small interval to predict
    - Valid limitation of model

## Conceptual Model: Step 1 (12)

Clustering

- NTAs used: <list>

- Docking Stations: Count

- Clusters: Count

- Map of Clusters by Color

## Conceptual Model: Step 2 (13)

Model Approaches

- 6 models attempted

## Conceptual Model: Step 2.1 (14)

- Plot of the availability by intervals to show the distribution

  - Mostly zero as the available number

## Model Results 1 (15)

- Certain input to model based on 3-hour intervals
- Results table


## Model Results 2 (16)

- Certain input to model based on 15-minute intervals
- Results table

## Model Results Visual (17)

- Plot of the preds and actuals for best scoring model

```{r}

```

## Prediction Function (18)

- Model prediction
  - Show how it works


## Current Achievements (19)

```{r}

```

## Future Works (20)

- Weather ... actually, can I predict weather? would that really work?
- Subway stations: Citi Bike offers valet 
- Model of all NYC
- Real-time clustering would be better

## Slide (21)

```{r}

```

## Slide (22)

```{r}

```