---
title: "Scratch Time Series Graphs"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

# ATTEMPT AT TIME SERIES GRAPHS USING FFP3 LIBRARY

```{r}
library(tidyverse)
library(ggplot2)
library(fpp3)
```

```{r}
# Let's create some other columns of value
# weekday
# day of the month
# trip duration in minutes
# start hour
citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = wday(ymd_hms(started_at),label=TRUE,abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual))

#citibike
```

Columns
- ride_id: Unique identifier of the ride, probably not meaningful to me
- rideable_type:

```{r}
nrow(citibike)
```

Checks out, that is the correct number

```{r}
head(citibike)
```

```{r}
citibike_ts <- citibike %>%
  mutate(start.ts = as_datetime(started_at)) %>%
  as_tsibble(index=start.ts, key = ride_id)
```

```{r}
citibike_ts[1:20,]
```
Hmm, I can't plot the data in this instance format, I need the info to follow a specific pattern interval before charting, so let's group by day and then chart that

```{r}
citibike_by_day <- citibike %>%
  mutate(day=as_date(started_at)) %>%
  group_by(day, member_casual) %>%
  summarize(count=n())

citibike_by_day
```


citibike %>%
  group_by(day, member_casual) %>%
  summarize(n=n(),
            weekday = weekday[1]) %>%
  group_by(weekday, member_casual) %>%
  summarize(n.m=mean(n)) %>%
  ggplot(aes(x=weekday, y=n.m)) + geom_bar(stat='identity') +
  labs(x = 'Day of Week',
       y = 'Number of Trips',
       title = 'Number of Bike Trips by Day of Week') +
  facet_wrap(~member_casual)

```{r}
citibike_by_day_ts <- citibike_by_day %>%
  as_tsibble(index=day, key=c(member_casual))

citibike_by_day_ts
```

```{r}
autoplot(citibike_by_day_ts, count) +
  labs(title = "Ride Trips by Day",
       subtitle = "Citi Bike NYC",
       y = "Ride Trips Counts")
```

```{r}
# Let's attempt hourly
# First convert started_at into a timestamp format
citibike$started_at <- as_datetime(as.character(citibike$started_at))

citibike
```
citibike_by_day




pedestrian %>%
  group_by_key() %>%
  index_by(Date_Time4 = ~ lubridate::floor_date(., "4 hour")) %>%
  summarise(Total_Count = sum(Count))


  
  citibike_by_day <- citibike %>%
  mutate(day=as_date(started_at)) %>%
  group_by(day, member_casual) %>%
  summarize(count=n())

citibike_by_day

```{r}
citibike_by_hour <- citibike %>%
  mutate(hour=lubridate::floor_date(started_at, "1 hour")) %>%
  group_by(hour, member_casual) %>%
  summarize(cnt=n())

citibike_by_hour
```

```{r}
citibike_by_hour_ts <- citibike_by_hour %>%
  as_tsibble(index=hour, key=c(member_casual))

citibike_by_hour_ts
```

```{r}
autoplot(citibike_by_hour_ts, cnt) +
  labs(title = "Ride Trips by Hour - Sept. 2022",
       subtitle = "Citi Bike NYC",
       y = "Ride Trips Counts")
```

Above is a good look at the time series display of trip counts per hour for the two user types

```{r}
citibike_by_hour_ts %>%
  gg_season(cnt, period = "week") +
  labs(y = "Ride Trips",
       title = "Seasonal plot: Weekly Trip Counts for Sept. 2022")
```

Rained all day on Tuesday, Sept 6 2022

```{r}
citibike_by_hour_ts %>%
  #filter(member_casual=='member') %>%
  gg_subseries(cnt, period = 'day') +
  labs(y = "Ride Trips",
       title = "Seasonal plot: Weekly Trip Counts for Sept. 2022")
```

Oh, this represents by hour across each day in which each section represents the full month. Still need to figure this out a bit.

```{r}
citibike_by_hour_ts %>%
  #filter(member_casual=='member') %>%
  gg_subseries(cnt, period = 'hour') +
  labs(y = "Ride Trips",
       title = "Seasonal plot: Weekly Trip Counts for Sept. 2022")
```

```{r}
citibike_by_hour_ts %>%
  ggplot(aes(x = hour, y = cnt)) +
  geom_point() +
  labs(x = "Temperature (degrees Celsius)",
       y = "Electricity demand (GW)")
```
Above not too valuable given I don't have a secondary value such as temperature, if I could get temperature or rainfall per hour, then perhaps there would be a linear-like relationship

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
