---
title: "Matrix of Start Stop by 15 minute intervals"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

# GOAL IS TO CREATE A MATRIX OF THE STARTS/ENDS FOR 15 MINUTE INTERVALS FOR EACH STATION

```{r}
# Required packages
library(tidyverse)
#library(ggplot2)
#library(skimr)
#library(tidyjson)
#library(geosphere)
library(lubridate)
# library(ggmap)
#library(forcats)
#library(scales)
```

```{r}
# Let's create some other columns of value
# weekday
# day of the month
# trip duration in minutes
# start hour
citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE)

#citibike
```

Columns
- ride_id: Unique identifier of the ride, probably not meaningful to me
- rideable_type:

```{r}
nrow(citibike)
```

 %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = wday(ymd_hms(started_at),label=TRUE,abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual))

```{r}
citibike
```

```{r}
# Ok, need to convert start time to actual datetime format
citibike$started_at <- as_datetime(as.character(citibike$started_at))
citibike$ended_at <- as_datetime(as.character(citibike$ended_at))
citibike$end_station_name[citibike$end_station_name == ''] <- "Abandoned"
citibike$end_station_id[citibike$end_station_id  == ''] <- "ABAN"

citibike
```

```{r}
# And now for grouping by 15 minutes

citibike_by_start.interval <- citibike %>%
  mutate(start.interval=lubridate::floor_date(started_at, "15 min")) %>%
  group_by(start_station_id, start.interval) %>%
  summarize(start.count=n())

citibike_by_start.interval
```

```{r}
citibike_by_end.interval <- citibike %>%
  mutate(end.interval=lubridate::floor_date(started_at, "15 min")) %>%
  group_by(end_station_id, end.interval) %>%
  summarize(count=n())

citibike_by_end.interval
```

```{r}
# Now I Need to merge by station_id and time interval

citibike_by_interval <- 
  merge(citibike_by_start.interval, 
        citibike_by_end.interval, 
        by.x=c('start_station_id','start.interval'), 
        by.y=c('end_station_id','end.interval'), all = TRUE)
```



```{r}
citibike_by_interval[is.na(citibike_by_interval)] <- 0
citibike_by_interval
```
```{r}
# Rename columns
colnames(citibike_by_interval)[1] <- "station_id"
colnames(citibike_by_interval)[2] <- "interval_start.time"
colnames(citibike_by_interval)[3] <- "depart.count"
colnames(citibike_by_interval)[4] <- "arrive.count"

citibike_by_interval
```

```{r}
# Calculate surplus
citibike_by_interval$surplus <- citibike_by_interval$arrive.count - citibike_by_interval$depart.count

citibike_by_interval
```

```{r}
citibike_by_interval_surplus <- citibike_by_interval %>%
  select(station_id, interval_start.time, surplus)

citibike_by_interval_surplus
```

```{r}

```

```{r}
mydateseq<-seq(as.POSIXct("2022-09-01"), by="15 min", length.out = 2880)  

# Output the data sequence to confirm the result are 15 minute intervals which start on the hour
mydateseq
```

```{r}
citibike_by_interval_surplus_wide <- citibike_by_interval_surplus %>%
  pivot_wider(names_from = station_id, values_from = surplus)

citibike_by_interval_surplus_wide[is.na(citibike_by_interval_surplus_wide)] <- 0


# For ordering
citibike_by_interval_surplus_wide <- citibike_by_interval_surplus_wide[order(citibike_by_interval_surplus_wide$interval_start.time),]
citibike_by_interval_surplus_wide
```
# Nice the above matrix is by time and then for all the columns are stations with the plus/minus for each 15 minute interval

# Next is to count the running total

```{r}
citibike_by_interval_surplus_wide
```

```{r}
citibike_by_interval_surplus_wide_cs <- cumsum(citibike_by_interval_surplus_wide[2:1659])
citibike_by_interval_surplus_wide_cs
```

```{r}
time_index <- citibike_by_interval_surplus_wide$interval_start.time
time_index
```

```{r}
citibike_by_interval_surplus_wide
```
```{r}
# Write the above to CSV
#write.csv(citibike_by_interval_surplus_wide, file = 'stations_sept22_interval_surplus.csv')
```

```{r}

citibike_by_interval_surplus_wide_cs <- cbind(time_index, citibike_by_interval_surplus_wide_cs)
citibike_by_interval_surplus_wide_cs
```

Alright, so `citibike_by_interval_surplus_wide_cs` is the running total of eash station

```{r}
tail(citibike_by_interval_surplus_wide_cs)
```
# So many of the stations are negative, which can't be possible. Zero is absolutely the lowest number of bikes

```{r}
ff = function(x)
{
    cs = cumsum(x)
    if (cs < 0) {
      cs = 0
    } else if (cs > 20) {
      cs = 20
    }
}
ff(citibike_by_interval_surplus_wide[2])
```

```{r}
cumsum.bounded <- function(x, lower.bound = 0, upper.bound = 500) {
    bsum <- function(x, y) min(upper.bound, max(lower.bound, x+y))
    if (length(x) > 1) Reduce(bsum, x, acc = TRUE) else x
}
```

```{r}
cumsum.bounded(citibike_by_interval_surplus_wide[2:1659], 0, 20)
```

```{r}

```

```{r}
library(dm)
```

```{r}
cumsum_bounded(citibike_by_interval_surplus_wide[2:1659], 0, 20)
```

Unfortunately I wasn't able to get the cumsum bounded ... yet, need to check the .cpp suggestion

https://stackoverflow.com/questions/21262424/bounded-cumulative-sum

https://stat.ethz.ch/pipermail/r-help/2010-November/261244.html

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
