---
title: "Scratch Station Info"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

# Focus on the station info

```{r}
# Required packages
library(tidyverse)
library(ggplot2)
library(skimr)
#library(tidyjson)
#library(geosphere)
library(lubridate)
# library(ggmap)
#library(forcats)
#library(scales)
library(jsonlite)
```

```{r}
# Read in trip data for September 2022
# Create variables for day of the month, start hour, weekday, trip duration, member_casual
citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = wday(ymd_hms(started_at),label=TRUE,abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual))
```

```{r}
#Station_Information.json
# Read in Citi bike Station information and only use the 'stations' object of the JSON
citibike_station_raw <- fromJSON("Station_Information.json")
citibike_station_info <- citibike_station_raw$data$stations

# Only keep needed columns
citibike_station_info <- citibike_station_info %>%
  select("station_type", "short_name", "has_kiosk", "region_id", "name", "lat", "station_id", "external_id", "capacity", "lon", "electric_bike_surcharge_waiver", "legacy_id")

# Output to confirm data appears reasonable
(citibike_station_info)
```

According to this file, there are 1708 stations

```{r}
n_distinct(citibike_station_info$name)
```

1708 unique names

```{r}
n_distinct(citibike_station_info$region_id)
```

3 region ids
According to specifications, a very generic identifier: https://github.com/MobilityData/gbfs/blob/master/gbfs.md#system_regionsjson

```{r}
citibike_station_info %>%
  filter(region_id == 71)
```

1627 stations in the NYC district (region_id==71)

```{r}
unique(citibike_station_info$region_id)
```


{
  "data": {
    "regions": [
      {
        "name": "JC District",
        "region_id": "70"
      },
      {
        "name": "NYC District",
        "region_id": "71"
      },
      {
        "name": "8D",
        "region_id": "158"
      },
      {
        "name": "Bronx",
        "region_id": "185"
      },
      {
        "name": "IC HQ",
        "region_id": "189"
      },
      {
        "name": "testzone",
        "region_id": "190"
      },
      {
        "name": "Hoboken District",
        "region_id": "311"
      }
    ]
  },
  "last_updated": 1667169676,
  "ttl": 5
}

Above confirms the 'NYC District' is region_id 71

```{r}
# Considering Time Series data
# This is directly from https://otexts.com/fpp3/appendix-using-r.html
# Load required packages
library(fpp3)

```


~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```