---
title: "Elevation and Distance v1"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

Scratch work to get the elevation for each station and distance between two stations

```{r}
# Required packages
library(tidyverse)
#library(ggplot2)
#library(skimr)
#library(tidyjson)
#library(geosphere)
#library(lubridate)
# library(ggmap)
#library(forcats)
#library(scales)
library(elevatr)
library(jsonlite)
library(rgdal)
```

```{r}
#Station_Information.json
# Read in Citi bike Station information and only use the 'stations' object of the JSON
citibike_station_raw <- fromJSON("Station_Information.json")
citibike_station_info <- citibike_station_raw$data$stations

# Only keep needed columns
citibike_station_info <- citibike_station_info %>%
  select("station_type", "short_name", "has_kiosk", "region_id", "name", "lat", "station_id", "external_id", "capacity", "lon", "electric_bike_surcharge_waiver", "legacy_id")

# Output to confirm data appears reasonable
(citibike_station_info)
```


```{r}
# Reordering columns to move lon and lat to first two columns, also dropped a few columns
col_order <- c('lon', 'lat', 'short_name', 'region_id', 'name', 'station_id', 'external_id', 'capacity')
citibike_station_info <- citibike_station_info[, col_order]
citibike_station_info
```

```{r}
# From: https://cran.r-project.org/web/packages/elevatr/vignettes/introduction_to_elevatr.html
prj_dd <- "EPSG:4326"

df_elev_epqs <- get_elev_point(citibike_station_info, prj = prj_dd, src = "epqs")
```

```{r}
(df_elev_epqs)
```

```{r}
write.csv(df_elev_epqs, file = 'stations_with_elevation.csv')
```

Do I need the distance between the stations? I don't think I really do besides for some EDA charts, but the availability is based on movement of bikes, via trip or via rebalancing, but the distance of the trips themselves wouldn't play a factor.

If I did want to get the distance, it seems https://openrouteservice.org/services/ and the API for time-distance matrix which is free with restrictions would be the best option to get info for free. otherwise, it seems the Google API would cost me over $1000, but again, I'm not sure what I'd do with that info.

Matrix API: max is 2500 combinations per request
Basic level API restrictions: Matrix* (500 / 40) 500 per day, 40 per minute


PERHAPS THE VALUE IN ELEVATION IS THE OVERALL NET SURPLUS IN A MONTH, IS THERE A STATION WITH A POSITIVE VALUE WHICH HAS A HIGHER ELEVATION?

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
