---
title: "Scratch Oct 31 Trip and Status v1"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

The only thing worthwhile in this document is the plot of one station that would capture the spikes of rebalancing and the inconsistency of that.

```{r libraries, echo=F, message=F, warning=F, include=F}
# Required packages
library(tidyverse)
library(ggplot2)
library(skimr)
library(lubridate)
library(jsonlite)
#library(MASS)
#library(fpp3)
#library(assertthat)
#library(igraph)
#library(ggraph)
#library(ggmap)
#library(leaflet)
#library(rgdal)
#library(RColorBrewer)
library(data.table)
```

```{r}
# Create additional columns of pertinence
# weekday, day of the month, trip duration in minutes, start hour
citibike_202210 <- fread("data/202210-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = lubridate::wday(ymd_hms(started_at), label=TRUE, abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual)) %>%
  rename(ride.id=ride_id, rideable.type=rideable_type, started.at=started_at,
         ended.at=ended_at, start.station.name=start_station_name, start.station.id=start_station_id,
         end.station.name=end_station_name, end.station.id=end_station_id, start.lat=start_lat,
         start.lng=start_lng, end.lat=end_lat, end.lng=end_lng, member.casual=member_casual)


# Figure out abandoned and label them
citibike_202210$end.station.name[citibike_202210$end.station.name == ''] <- "Abandoned"
citibike_202210$end.station.id[citibike_202210$end.station.id  == ''] <- "ABAN"

# Output DF if needed
head(citibike_202210)
```

```{r}
citibike_oct31 <- citibike_202210 %>%
  filter(started.at >= '2022-10-30 20:00:00') # Minus 4 hours to get midnight of Oct. 31
```

```{r}
citibike_oct31 <- citibike_oct31[order(citibike_oct31$started.at),]
head(citibike_oct31)
```

```{r}
citibike_oct31 %>%
  filter(start.station.id == "3960.01") %>%
  summarize(min(ended.at))
```

Apparently no activity before 5am on Oct. 31

# Building table of surplus counts

```{r calculate.surplus, echo=F, message=F, warning=F, include=T}
# Create table of start to end station IDs
stations_cols <- citibike_oct31 %>%
  dplyr::select(start.station.id, end.station.id)

stations_table <- as.data.frame((table(stations_cols)))

stations_table <- stations_table %>%
  filter(Freq > 0)

stations_table_order <- stations_table[order(-stations_table$Freq),]

stations_table_dif <- stations_table_order %>%
  filter(as.character(start.station.id) != as.character(end.station.id))

# Create table of start to end station IDs
stations_cols <- citibike_oct31 %>%
  dplyr::select(start.station.id, end.station.id)
stations_table <- as.data.frame((table(stations_cols)))

stations_table <- stations_table %>%
  filter(Freq > 0)

stations_table_order <- stations_table[order(-stations_table$Freq),]

stations_table_dif <- stations_table_order %>%
  filter(as.character(start.station.id) != as.character(end.station.id))

stations_table_dif1 <- stations_table_dif
stations_table_dif2 <- stations_table_dif

# Added all=TRUE to account for one-sided counts
stations_table_dif_merge <- 
  merge(stations_table_dif1, 
      stations_table_dif1, 
      by.x=c('start.station.id','end.station.id'), 
      by.y=c('end.station.id','start.station.id'), all = TRUE)

# Set the NA (one-sided trips) to count of 0
stations_table_dif_merge[is.na(stations_table_dif_merge)] <- 0

stations_table_dif_merge$surplus <- stations_table_dif_merge$Freq.y - stations_table_dif_merge$Freq.x
stations_table_dif_merge <- stations_table_dif_merge[order(-stations_table_dif_merge$surplus),]

# Remove rows with start station id equal to ABAN for abandoned, those are a result of the merge all=TRUE and should not exist
stations_table_dif_merge <- stations_table_dif_merge %>% filter(start.station.id != 'ABAN')

# Want to identify the surplus (or not) by station for the month
station_surplus_count <- stations_table_dif_merge %>%
  group_by(start.station.id) %>%
  summarize(surplus.sum=sum(surplus))


station_surplus_count <- station_surplus_count[order(-station_surplus_count$surplus.sum),]
colnames(station_surplus_count)[1] <- "station.id"
```




```{r}
station_surplus_count <- station_surplus_count[order(station_surplus_count$surplus.sum),]
```

```{r}
station_surplus_count
```

So, over the course of 1 day, Oct. 31, 2022, these 10 stations have at least -30 or more shortage of bikes.

```{r}
citibike_oct31 %>%
  filter(start.station.id == "5329.03")
```

```{r}
citibike_bike_avail <- fread("bike_avail_by_station_and_time.csv", data.table=FALSE, check.names=FALSE)
```

```{r}
citibike_bike_avail_426 <- citibike_bike_avail %>%
  dplyr::select("timestamp", "426") %>%
  filter(timestamp < "2022-10-31 20:00:00")
citibike_bike_avail_426
```

```{r}
station_boro <- fread("stations_with_boro_and_hood.csv", data.table=FALSE, check.names=FALSE)
```

```{r}
station_boro %>%
  filter(ntacode == "BK61")
```

```{r}
# And now for grouping by 15 minutes
citibike_by_start.interval <- citibike_oct31 %>%
  filter(start.station.id == "5329.03") %>%
  mutate(start.interval=lubridate::ceiling_date(started.at, "15 min")) %>%
  group_by(start.station.id, start.interval) %>%
  summarize(start.count=n())

(citibike_by_start.interval)
```

```{r}
citibike_by_end.interval <- citibike_oct31 %>%
  filter(end.station.id == "5329.03") %>%
  mutate(end.interval=lubridate::ceiling_date(ended.at, "15 min")) %>%
  group_by(end.station.id, end.interval) %>%
  summarize(count=n())

(citibike_by_end.interval)
```

```{r}
# Now I Need to merge by station_id and time interval

citibike_by_interval <- 
  merge(citibike_by_start.interval, 
        citibike_by_end.interval, 
        by.x=c('start.station.id','start.interval'), 
        by.y=c('end.station.id','end.interval'), all = TRUE)
```

```{r}
citibike_by_interval[is.na(citibike_by_interval)] <- 0
citibike_by_interval
```

```{r}
# Rename columns
colnames(citibike_by_interval)[1] <- "station_id"
colnames(citibike_by_interval)[2] <- "interval_start.time"
colnames(citibike_by_interval)[3] <- "depart.count"
colnames(citibike_by_interval)[4] <- "arrive.count"

(citibike_by_interval)
```

```{r}
# Calculate surplus
citibike_by_interval$surplus <- citibike_by_interval$arrive.count - citibike_by_interval$depart.count

(citibike_by_interval)
```

```{r}
citibike_by_interval_surplus <- citibike_by_interval %>%
  dplyr::select(station_id, interval_start.time, surplus) %>%
  rename(interval.start.time = interval_start.time)

(citibike_by_interval_surplus)
```

```{r}
citibike_bike_avail_426
```

```{r}

citibike_station_426 <- 
  merge(citibike_by_interval_surplus, 
        citibike_bike_avail_426, 
        by.x=c('interval.start.time'), 
        by.y=c('timestamp'), all = TRUE)
```

```{r}
citibike_station_426 <- citibike_station_426 %>%
  dplyr::select(interval.start.time, surplus, "426") %>%
  rename(bikes.avail="426")

citibike_station_426
```

```{r}
citibike_station_426[is.na(citibike_station_426)] <- 0
citibike_station_426 <- head(citibike_station_426,-2)
citibike_station_426
```

```{r}
citibike_bike_avail <- citibike_bike_avail %>%
  dplyr::select(-V1)
```

```{r}
citibike_bike_avail
```

```{r}
citibike_bike_avail_long <- citibike_bike_avail %>%
  pivot_longer(!timestamp, names_to = "station.id", values_to = "bikes.avail")

citibike_bike_avail_long
```

```{r}
citibike_bike_avail_long_trim <- citibike_bike_avail_long %>%
  filter(as.integer(station.id) == 3582)
```

```{r}
p<-ggplot(citibike_bike_avail_long_trim, aes(x=timestamp, y=bikes.avail, group=station.id)) +
  geom_line(aes(color=station.id))+
  geom_point(aes(color=station.id))
p
```

The above shows there is rebalancing at this station but it's not consistent.

```{r}
citibike_bike_avail_long_trim
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```