---
title: "DATA 698 Project v1"
subtitle: "Predicting Citi Bike Availability in NYC"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    toc: true
    theme: "Boadilla"
    colortheme: "default"
    fonttheme: "structurebold"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Problem and Objective 

Introduce yourselves and describe your problem.
Explain your objectives, challenges of your work, proposed methodologies, and the assumptions you made while conducting modeling and/or analysis.
Provide an overview of your approach and/or conceptual model (please do not present your code directly). 
Describe the results you obtain and summarize the current achievements and possibility of future works.

## Challenges and Assumptions

- Too much data
  - Over 3.5 million trips in month of Sept. 2022
- Rebalancing identification
- User friendly approach
  - Inputs to output

## Proposed Methodologies

- Inputs
  - Latitude and longitude
  - Day of the Week
  - Time of Day
  - Number of Bikes 
- Citi Bike offers live map of availability
- Lyft provides real-time availability

- Time series model
- Poisson distribution and Negative Binomial given the over-dispersion

## Overview of Approach

- Data from API call every 15 minutes for two weeks
  - Citi bike availability at each station
  - Two weeks is small interval to predict
    - Valid limitation of model

## EDA 1

```{r libraries, echo=F, message=F, warning=F, include=F}
# Required packages
library(tidyverse)
library(ggplot2)
library(skimr)
library(lubridate)
library(fpp3)
library(assertthat)
library(igraph)
library(ggraph)
library(ggmap)
library(leaflet)
library(rgdal)
library(RColorBrewer)
library(jpeg)
library(data.table)
```

```{r read.trip.data, echo=F, message=F, warning=F, include=F, eval=T}
# Create additional columns of pertinence
# weekday, day of the month, trip duration in minutes, start hour
citibike <- fread("data/202209-citibike-tripdata.csv", data.table=FALSE, check.names=TRUE) %>%
#citibike <- read.csv("data/202209-citibike-tripdata.csv", check.names=TRUE) %>%
  mutate(day = factor(mday(ymd_hms(started_at))),
         start.hour=factor(hour(ymd_hms(started_at))),
         weekday = lubridate::wday(ymd_hms(started_at), label=TRUE, abbr=TRUE),
         trip.duration = as.numeric(difftime(ended_at,started_at,units="mins")),
         member_casual = factor(member_casual)) %>%
  rename(ride.id=ride_id, rideable.type=rideable_type, started.at=started_at,
         ended.at=ended_at, start.station.name=start_station_name, start.station.id=start_station_id,
         end.station.name=end_station_name, end.station.id=end_station_id, start.lat=start_lat,
         start.lng=start_lng, end.lat=end_lat, end.lng=end_lng, member.casual=member_casual)

# Figure out abandoned and label them
citibike$end.station.name[citibike$end.station.name == ''] <- "Abandoned"
citibike$end.station.id[citibike$end.station.id  == ''] <- "ABAN"

# Output DF if needed
head(citibike)
```

```{r trip.by.hour.time.series, echo=F, message=F, warning=F, include=T}

citibike$started.at.ts <- as_datetime(as.character(citibike$started.at))

citibike_by_hour <- citibike %>%
  mutate(hour=lubridate::floor_date(started.at.ts, "1 hour")) %>%
  group_by(hour, member.casual) %>%
  summarize(cnt=n())
#citibike_by_hour

citibike_by_hour_ts <- citibike_by_hour %>%
  as_tsibble(index=hour, key=c(member.casual))
#citibike_by_hour_ts

autoplot(citibike_by_hour_ts, cnt) +
  labs(title = "Ride Trips by Hour - Sept. 2022",
       subtitle = "Citi Bike NYC",
       y = "Ride Trips Counts")
```

## EDA 2

```{r trip.by.hour.seasonal, echo=F, message=F, warning=F, include=T}
citibike_by_hour_ts %>%
  gg_season(cnt, period = "week") +
  labs(y = "Ride Trips",
       title = "Seasonal plot: Weekly Trip Counts for Sept. 2022")
```

## EDA 3

```{r display.surplus.by.station, echo=F, message=F, warning=F, include=T}
img <- readJPEG("stations_surplus_sum.jpg")
plot(1:10,ty="n", axes = 0, xlab='', ylab='', main='Overall Monthly Surplus by Docking Station')
rasterImage(img,-1,-1,12,12)
```

## focus on Crown Heights

```{r}

```

## Conceptual Model: Step 1

Clustering

## Conceptual Model: Step 2

Modeling

## Model Results

- Results table

## Slide 10

```{r}

```

## Current Achievements

- Clustering
- Model prediction


## Slide 12

```{r}

```

## Future Works

- Weather ... actually, can I predict weather? would that really work?
- Subway stations: Citi Bike offers valet 
- Model of all NYC
- Real-time clustering would be better

## Slide 14

```{r}

```

## Slide 15

```{r}

```