---
title: "Citibike Station Availability"
author: "Philip Tanofsky"
date: "`r Sys.Date()`"
output: html_document
---

At 8:30p 11/11/22, goal is to 1) read in one of the station availability API responses from Chron script and 2) determine the borough and neighborhood for each station based on lat/long


```{r}
# Required packages
library(tidyverse)
library(ggplot2)
library(skimr)
#library(tidyjson)
#library(geosphere)
library(lubridate)
# library(ggmap)
#library(forcats)
#library(scales)
library(geojsonR)
#library(rjson)
library(jsonlite)
#library(sp)

library(sp)
library(rgdal)
library(maps)
library(dplyr)
```

GeoJson file: nyc-neighborhoods_gis.geojson

Station API response in JSON: cb_station_2022-10-30_21_16_00.json


```{r}

# https://data.cityofnewyork.us/City-Government/2010-Neighborhood-Tabulation-Areas-NTAs-/cpf4-rkhq
#nyc_hoods_js = FROM_GeoJson("data/nyc_nta_2010.geojson")
nyc_hoods_js = readOGR("data/nyc_nta_2010.geojson")

nyc_hoods_js
```

```{r}
cb_station_json <- jsonlite::read_json('data/cb_station_2022-10-30_21_16_00.json', simplifyVector = TRUE)

stations_avail <- cb_station_json$data$stations
```

```{r}
stations_avail
```

```{r}
#Station_Information.json
# Read in Citi bike Station information and only use the 'stations' object of the JSON
citibike_station_raw <- jsonlite::fromJSON("Station_Information.json")
citibike_station_info <- citibike_station_raw$data$stations

# Only keep needed columns
citibike_station_info <- citibike_station_info %>%
  select("station_type", "short_name", "region_id", "name", "lat", "station_id", "capacity", "lon", "legacy_id")

# Output to confirm data appears reasonable
(citibike_station_info)
```

```{r}
```


```{r}
# https://datawanderings.com/2018/09/01/r-point-in-polygon-a-mathematical-cookie-cutter/
nyc_hoods_js@proj4string
```

```{r}
coordinates(citibike_station_info) <- c("lon","lat")
as(citibike_station_info,"SpatialPoints")
```

```{r}
proj4string(citibike_station_info) <- CRS("+proj=longlat +datum=WGS84")
proj4string(citibike_station_info) <- proj4string(nyc_hoods_js)
```

```{r}
pointsinpoly <- over(citibike_station_info, nyc_hoods_js)

pointsinpoly
```

```{r}
output <- cbind(citibike_station_info,pointsinpoly)
output
```

Ugh, okay, I was able to determine the neighborhood and borough for each of the 1708 stations

# Ok, now for the availability

```{r}
stations_avail %>%
  filter(station_status == 'active' & num_bikes_available >= 5)
```

```{r}
cb_stations_avail <- 
  merge(stations_avail, 
      output, 
      by.x=c('legacy_id'), 
      by.y=c('legacy_id'), all.x = TRUE)

#res[is.na(res)] <- 0
time_station_avail <- cb_stations_avail %>%
  select(legacy_id, station_status, num_bikes_available, short_name, name, capacity, ntaname, boro_name, lon, lat)

time_station_avail
```

```{r}
time_station_avail %>%
  filter(station_status == 'active' & boro_name == "Brooklyn" & num_bikes_available < 5)
```

```{r}

```

```{r}

```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

~~~~~

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
